---
title: "Synthetic campaign pipeline simulation"
output: html_notebook
---

```{r setup, include=FALSE}
# Show code chunks
knitr::opts_chunk$set(echo = FALSE)
```

Suppose your institution is planning a campaign and leadership wants to know how risky it is to only have one naming-gift-level donor in the solicitation pipeline.

Create some synthetic data under four different scenarios: high variance with a single very large gift and multiple small ones, followed by medium, low, and no variance pipelines where the range of solicitations becomes smaller and smaller.

```{r, warning=FALSE, message=FALSE}
# Load packages
library(ggplot2)
library(tidyverse)
```
```{r}
# Function to format as dollars (in millions)
dollars <- function(nbr) {
  paste0('$', nbr %>% formatC(format = 'f', big.mark = ',', digits = 0), 'M')
}

# Generate pipeline data (in millions)
# Note that e.g. rep(1, 6) is shorthand for c(1, 1, 1, 1, 1, 1)
sol_hivar <- c(rep(1, 6), rep(2, 3), rep(5, 9), 13, 80)
sol_medvar <- c(rep(2, 5), rep(5, 9), rep(10, 4), 15, 40)
sol_lowvar <- c(rep(5, 8), rep(6, 4), rep(10, 7), 16)
sol_novar <- rep(7.5, 20)

# Check that they all have the same length and sum
if(length(sol_hivar) != length(sol_medvar) | length(sol_hivar) != length(sol_lowvar) | length(sol_hivar) != length(sol_novar)) {
  print('ERR: different lengths')
}
if(sum(sol_hivar) != sum(sol_medvar) | sum(sol_hivar) != sum(sol_lowvar) | sum(sol_hivar) != sum(sol_novar)) {
  print('ERR: different sums')
}
```

```{r}
# Summary plot
data.frame(
  hivar = sol_hivar
  , medvar = sol_medvar
  , lowvar = sol_lowvar
  , novar = sol_novar
) %>% gather(
  "pipeline", "dollars", 1:4
) %>% mutate(
  pipeline = factor(pipeline) %>% forcats::fct_inorder()
) %>%
  ggplot(aes(x = dollars, fill = pipeline)) +
  geom_histogram(binwidth = 1) +
  facet_grid(pipeline ~ .) +
  scale_x_continuous(labels = scales::dollar) +
  labs(x = "dollars (millions)", y = "count of solicitations at this level")
```

This represents 4 different campaigns, with MG pipelines of `r length(sol_hivar) %>% I()` gifts summing up to `r sum(sol_hivar) %>% dollars() %>% I()` in asks. Note that all synthetic pipelines have the same number of proposals and the same mean:

```{r}
stats <- function(sols) {
  data.frame(
    n = length(sols)
    , StDev = sd(sols)
    , min = min(sols)
    , median = median(sols)
    , mean = mean(sols)
    , max = max(sols)
  )
}

rbind(
  hivar = stats(sol_hivar)
  , medvar = stats(sol_medvar)
  , lowvar = stats(sol_lowvar)
  , novar = stats(sol_novar)
)
```

Let's run a Monte Carlo simulation to see what outcomes might be expected for each of these campaigns.

```{r}
# Generate probability vectors for Monte Carlo style simulation
p <- 1/3
iter <- 2500
set.seed(201494)
pmat <- matrix(
  runif(length(sol_hivar) * iter)
  , nrow = iter
  , ncol = length(sol_hivar)
)
pmat <- ifelse(pmat <= p, 1, 0)

# Construct data frame of simulation results
sol_sims <- data.frame(
  hivar = pmat %*% sol_hivar
  , medvar = pmat %*% sol_medvar
  , lowvar = pmat %*% sol_lowvar
  , novar = pmat %*% sol_novar
) %>% gather(
  "pipeline", "dollars", 1:4
) %>% mutate(
  pipeline = factor(pipeline) %>% forcats::fct_inorder()
)
```

Assume all proposals have a $p =$ `r round(p, digits = 3) %>% I()` chance of closing at the expected value, and the probabilities are all independent, so that one proposal closing or not doesn't affect the next.

Here are results for `r iter %>% I()` simulations of each campaign under those assumptions.

```{r}
rbind(
  hivar = stats(pmat %*% sol_hivar)
  , medvar = stats(pmat %*% sol_medvar)
  , lowvar = stats(pmat %*% sol_lowvar)
  , novar = stats(pmat %*% sol_novar)
)
```

Plotting the results of the simulation:


```{r, message = FALSE, warning = FALSE, error = FALSE}
# Average result across pipeline simulations
pipeavg <- function(pipeline_name) {
  sol_sims %>% filter(pipeline == pipeline_name) %>% select(dollars) %>% unlist() %>% mean()
}

# Plot results
sol_sims %>%
  ggplot(aes(x = dollars, color = pipeline)) +
  geom_density() +
  geom_vline(xintercept = pipeavg('hivar'), linetype = 'dashed', color = 'red') +
  geom_vline(xintercept = pipeavg('medvar'), linetype = 'dashed', color = 'green') +
  geom_vline(xintercept = pipeavg('lowvar'), linetype = 'dashed', color = 'cyan') +
  geom_vline(xintercept = pipeavg('novar'), linetype = 'dashed', color = 'purple') +
  scale_x_continuous(labels = scales::dollar) +
  labs(x = 'dollars (millions)', y = 'probability density')
```

Though all campaigns are expected to raise `r {sum(sol_hivar) * p} %>% dollars() %>% I()` on average, the success of the high-variance pipeline campaign depends almost entirely on the single `r max(sol_hivar) %>% dollars() %>% I()` proposal. Specifically, in any given simulation there is a `r {(1 - round(p, digits = 3)) * 100} %>% paste0(., '%') %>% I()` chance this gift does not come in, and when it doesn't, making the `r {sum(sol_hivar) * p} %>% dollars() %>% I()` goal is very unlikely. In fact, out of the `r nrow(pmat[which(pmat[, 20] == 0), ]) %>% I()` simulations where the `r max(sol_hivar) %>% dollars() %>% I()` gift did not close, only `r sum(pmat[which(pmat[, 20] == 0), ] %*% sol_hivar >= {sum(sol_hivar) * p}) %>% I()` exceeded the `r {sum(sol_hivar) * p} %>% dollars() %>% I()` expected value (mean).

In the high-variance situation, reporting that the expected campaign outcome is `r {sum(sol_hivar) * p} %>% dollars() %>% I()` *is nearly meaningless!* High variance implies low confidence.

## Session Info

```{r}
devtools::session_info()$platform
```